Concurrent Programming in Harmony

### 2021-12-25 book/paper.tex
Latest

Harmony in fact supports this.
It has an operator \textbf{countLabel}~$L$,
\index{countLabel operator}%
where $L$ is the name of the label (in this case, <{cs}>).
The operator returns the number of threads executing at that label.
Method \textbf{countLabel} only exists for specification purposes---do not
use it in normal code.
If you run the code through Harmony, the assertion should fail because
there is no code yet for safely entering and exiting the critical section.

@@

The two things to note here is the use of the \textbf{choose}
expression and the <{assert}> statement.  Using <{choose}>,
we are able to express searching through all possible strings of
references without a complicated nested iteration.
Using \texttt{assert}, we are able to express the anomaly we are
looking for.

@@

Peterson's algorithm relies on a \emph{sequential consistent memory model}
and hence the \textbf{sequential} statement: without it Harmony will
complain about data races.
More precisely, the \textbf{sequential} statement says that the
program relies on memory load and store instructions operating on
the indicated variables to be performed sequentially, and that this
order should be consistent with the order of operations invoked on each
thread.
The default memory models of C and Java are not sequentially
consistent.
The unfortunately named \texttt{volatile} keyword in Java has
a similar function as Harmony's \texttt{sequential} keyword.
Like many constructions in Java, its \texttt{volatile} keyword
was borrowed from C and C++.
However, in C and C++, they do \emph{not} provide sequential
consistency, and one cannot implement Peterson's algorithm in
C or C++ directly.

@@

Harmony does not report data races in two cases.
First, using the \textbf{sequential} statement, you can specify that
concurrent access to the specified variables is intended.
Second, if the accesses are within an atomic statement block, then they
are not considered part of a data race.

@@

\begin{comment}
For the same reason that you should not use atomic
statements in your ordinary synchronization code, you should not use
\textbf{sequential} statements either.  Neither is supported by regular
programming languages.  They exist in Harmony for the purpose of
defining atomic instructions and for testing.
The statements should only appear in modules that define synchronization
primitives or in test programs.
\end{comment}

@@

This causes the context of the thread to be saved in <{s}> and
the thread to be no longer running.
Another thread can revive the thread using the \textbf{go}
\index{go statement}%
statement:

@@

Here <{c}> is a context and <{r}> is a Harmony value.
It causes a thread with context <{c}> to be added to the state that has
just executed the \textbf{stop}
\index{stop expression}%
expression.  The \textbf{stop} expression returns the value <{r}>.

\autoref{fig:suspension} shows the lock interface using suspension.
It is implemented as follows:
\begin{itemize}
\item A lock maintains both a boolean indicating whether the
lock is currently acquired and a list of contexts of threads that want to
acquire the lock.
\item
<{acquire()}>
\index{acquire}%
acquires the lock if available and suspends the invoking thread if not.
In the latter case, the context of the thread is added to the end of the list of contexts.
Note that \textbf{stop} is called within an atomic statement block---this is
the only exception to such an atomic statement block running to completion.
While the
thread is running no other threads can run, but when the thread suspends itself
other threads can run.
\item
<{release()}>
\index{release}%
checks to see if any threads are waiting to acquire the lock.
If so, it uses the <{head}> and <{tail}>
methods from the \texttt{list} module (see \autoref{ap:module})
to resume the first thread that got
suspended and to remove its context from the list.
\end{itemize}
Selecting the first thread is a design choice.  Another implementation could
have picked the last one, and yet another implementation could have used
<{choose}> to pick an arbitrary one.  Selecting the first is a common
choice in lock implementations as it prevents starvation.

@@

The reason is that \texttt{queue.get()} changes the state by
acquiring a lock, but the expressions in <{assert}>
statements (or \textbf{invariant} statements)
are not allowed to change the state.

@@

\autoref{fig:qtest1} shows a program that
checks concurrent <{put}> operations using
<{N}> threads.  Thread \texttt{putter(<{v}>)}
adds <{v}> to the queue.
Thread \texttt{main()} waits until all \texttt{putter}
threads have finished, and then dequeues <{N}>
items, verifying that they form the set
$\{ 0 .. \mathtt{N} - 1 \}$.
% Note that it is perfectly fine to use \textbf{sequential}
% and \textbf{atomic} statements in test code---one should
% only avoid those in production code.

@@

While having five different test programs is fine,
it would be nice if we can combine them all in one
test program.
This has to be done with some care, as running the
concurrent test programs simultaneously would
lead to a state space explosion.
Instead, we can leverage the \textbf{choose}
expression.
\autoref{fig:multitest} shows how one might go
about this.
Harmony will run all the tests, but the
tests do not interfere with one another.

@@

On the left is a Hoare monitor.
When some thread <{p}> is in the bathroom, it can invoke \textbf{signal}
on some non-empty bedroom associated with a condition that holds.
This select a thread <{q}> in that bedroom.
As a result of invoking \textbf{signal}, <{p}> goes back into the hall
and <{q}> enters the bathroom.\footnote{Per Brinch Hansen' monitors have
a \textbf{continue} operation instead of \textbf{signal}.  The difference
is that threads that invoke \textbf{continue} leave the monitor rather
than returning to the hall.}
Importantly, when <{q}> enters the bathroom, the condition that <{q}> was
waiting for is guaranteed to hold.

@@

\autoref{fig:actor} illustrates the actor approach.
There are three client threads that each want to be assigned
a unique identifier from the set $\{ 0, 1, 2 \}$.
Normally one would use a shared 0-initialized counter and a lock.
Each client would acquire the lock, get the value of the counter
and increment it, and release the lock.
Instead, in the actor approach the counter is managed by a
separate server thread.
The server never terminates, so it is spawned with the keyword
\textbf{eternal} to suppress non-terminating state warnings.
Each client sends a request to the server, consisting in this case
of simply the queue to which the server must send the response.
The server maintains a local, zero-initialized counter variable.
Upon receiving a request, it returns a response with the value of
the counter and increments the counter.  No lock is required.

@@

\autoref{fig:leader} describes the protocol and its test cases in Harmony.
In Harmony, processors can be modeled by threads and there are a variety of ways in
which one can model a network using shared variables.
Here, we model the network as a set of messages.
The \texttt{send} method atomically adds a message to this set.
Messages are tuples with three fields:
(<{dst}>, <{id}>, <{found}>).
<{dst}> is the identifier of the destination processor;
<{id}> is the identifier that is being forwarded; and
<{found}> is a boolean indicating the second phase of the protocol.
The \texttt{receive}(<{self}>) method looks for all messages
destined for the processor with identifier <{self}>.

@@

Method \texttt{processor}(<{self}>, <{succ}>) is the
code for a processor with identifier <{self}> and successor
<{succ}>.
It starts simply by sending its own identifier to its successor.
The processor then loops until it discovers the identifier of the
leader in the second phase of the protocol.
A processor waits for a message using the Harmony
\textbf{atomically when exists} statement.
This statement takes the form <{atomically when exists v in s}>: \emph{statement block}
where <{s}> is a set and <{v}> is variable that is bound to an element of <{s}>.
The properties of the statement are as follows:
\begin{itemize}
\item it waits until <{s}> is non-empty;
\item it is executed atomically;
\item <{v}> is selected non-deterministically, like in the <{choose}> operator.
\end{itemize}

@@

\autoref{fig:2pc2} contains the code for transfers and inquiries, as well
as tests.
The <{receive()}> method is used by coordinators in an
\textbf{atomically when exists}
statement to wait for a response from each bank involved in a transaction.
Argument \texttt{self} is the
identifier of the coordinator and \texttt{sources} is the set of banks.
It returns the empty set if there not yet responses from all banks.
Otherwise it returns a singleton set containing the set of responses,
one for each source.

@@

\begin{comment}
Note the use of the \textbf{possibly} statement in this code.
It is similar to <{assert}>, but where \texttt{assert} checks that
the predicate holds in \emph{all} executions, \texttt{possibly} checks
that the predicate holds in at least one execution.
The purpose is to check that the code doesn't simply always abort---there
exist executions where the code commits.
In practice, hopefully \emph{most} executions commit, but Harmony is not
capable of probabilistic reasoning.
\end{comment}

@@

In Line~22, a processor waits for $<{N}> - <{F}>$ messages using
the Harmony \textbf{atomically when exists} statement.
Since Harmony has to check all possible executions of the protocol, the
\texttt{receive}(<{round}>, <{k}>) method returns all \emph{subbags} of messages
for the given round that have size $k = <{N}> - <{F}>$.  The method uses a
dictionary comprehension to filter out all messages for the given <{round}>
and then uses the \texttt{bag.combinations} method to find all combinations of size <{k}>.
The \textbf{atomically when exists} statement waits until there is at least one such combination and
then chooses an element, which is bound to the <{quorum}> variable.
The body of the statement is then executed atomically.
This is usually how distributed algorithms are modeled, because they can only interact
through the network.  There is no need to interleave the different processes other
than when messages are delivered.  By executing the body atomically, a lot of
unnecessary interleavings are avoided and this reduces the state space that must
be explored by the model checker significantly.

The body of the \textbf{atomically when exists} statement contains the core of the algorithm.
Note that $<{N}> - <{F}> = 2<{F}> + 1$, so that the number of
messages is guaranteed to be odd.  Also, because there are only 0 and 1 values, there
must exist a majority of zeroes or ones.  Variable <{count[0]}> stores the number
of zeroes and <{count[1]}> stores the number of ones received in the round.
The rules of the algorithm are simple:
\begin{itemize}
\item update <{estimate}> to be the majority value;
\item if the quorum is unanimous, decide the value.
\end{itemize}
After that, proceed with the next round.

@@

A statement can be preceded by the <{atomically}> keyword to
make the statement atomic.
The statement <{atomically: S1; S2; ...}> evaluates statements
<{S1, S2, ...}> atomically.
This means that the statement runs indivisibly---no other thread can
interleave in the atomic statement.  The only exception
to this is if the atomic block executes a \textbf{stop} expression.
In this case, another thread can run.  When the original thread
is resumed (using a \textbf{go} statement), it is once again
atomically executing.

@@

The statement <{await b}> is equivalent to \textbf{when <{b}>: pass}.
It is intended to improve readability of your code.

@@

\subsection*{\textbf{for ... in ... [where ...]}}

@@

\subsection*{\textbf{from ... import}}

The statement \textbf{from} \texttt{x} \textbf{import} \texttt{a, b, ...}
imports module \texttt{x} and makes its constants $a, b, ...$ also constants
in the current module.  If a module is imported more than once, its code
is only included the first time.  The constants will typically be the
names of methods (program counter constants) within the module.

You can import all constants from a module <{m}>
(including program counter constants) using the statement
\textbf{from} <{m}> \textbf{import} \texttt{*}.
This, however, excludes constants whose names start with the character
\texttt{\_}: those are considered \emph{private} to the module.

\subsection*{\textbf{go}}

The statement \textbf{go} <{c e}> starts a thread with context <{c}> that
has executed a \textbf{stop} expression.  The \textbf{stop} expression
returns value <{e}>.
The same context can be started multiple times, allowing threads to
\emph{fork}.

\subsection*{\textbf{if ... [elif ...]* [else]}}

Harmony supports <{if}> statements.
In its most basic form, <{if}> $c: S_1; S_2; ...$ evaluates <{c}>
and executes statements $S_1, S_2, ...$ if and only if boolean expression
<{c}> evaluated to true.
Harmony checks that <{c}> is either <{False}> or <{True}>---if neither
is the case, Harmony reports an error.
The statement <{if}> $c: S_1, S_2, ...$ <{else}>:
$T_1; T_2; ...$ is similar, but executes statements $T_1, T_2, ...$
if and only if <{c}> evaluated to false.
You can think of \textbf{elif} $c:$ as shorthand for
<{else}>: <{if}> $c:$.

\subsection*{\textbf{import}}

The statement \textbf{import} $\mathtt{m}_1, \mathtt{m}_2, ...$ imports modules
$\mathtt{m}_1, \mathtt{m}_2, ...$ in that order.
If a module is imported more than once,
its code is only included the first time.
The constants (including method constants) and shared variables declared 
in that module can subsequently be referenced by prepending ``<{m}>.''.
For example, method <{f()}> in imported module <{m}> is invoked
by calling \texttt{m.f()}.  If you would prefer to invoke it simply as
<{f()}>, then you have to import using the statement
\textbf{from} <{m}> \textbf{import} \texttt{f}.

\subsection*{\textbf{invariant}}

The statement \textbf{invariant} <{c}> declares that boolean expression <{c}>
is an invariant.  <{c}> is only allowed to read shared variables and
is evaluated atomically after every state change.
If it ever evaluates to <{False}> Harmony reports and error.  Harmony
also reports an error if the expression evaluates to a value other
than <{False}> or <{True}>.

@@

You can introduce new bound variables in a method using the
<{let}> statement.
The statement <{let}> $a = b: S_1; S_2, ...$ evaluates <{b}>, assigns
the result to read-only variable <{a}>, and evaluates statements
$S_1, S_2, ...$.  <{let}> supports pattern matching, so you can
write <{let}> $x, (y, z) = b: S_1; S_2, ...$.  This will only work
if <{b}> is a tuple with two elements, the second of which also being a
tuple with two elements---if not, Harmony will report an error.

<{let}> statements may be nested, such as
<{let}> $a_1 = b_1$ <{let}> $a_2 = b_2: S_1; S_2; ...$.
Doing so can improve readability by reducing indentation
compared to writing them as separate statements.
Compare the following two examples:

@@

\subsection*{\textbf{pass}}

The \textbf{pass} statement does nothing.

\subsection*{\textbf{print}}

The statement \textbf{print <{e}>} evaluates <{e}>
and adds the result to the print log.  The print log is used to
create an ``external behavior DFA'' for the Harmony program.

\subsection*{\textbf{sequential}}

In Harmony, shared variable \texttt{Load} and \texttt{Store} operations
are atomic and have \emph{sequential consistency}.
However, Harmony does check for \emph{data races}.  A data race occurs
when two or more threads simultaneously access the same shared variable,
with at least one of the accesses being a \texttt{Store} operation
\emph{outside of an atomic block}.
If so, Harmony will report an error.  This error can be suppressed by
declaring the shared variable as sequential.  In particular, the statement
\texttt{sequential} $x, y, ...$ specifies that the algorithm assumes
that the given variables have sequential consistency.

@@

\subsection*{\textbf{spawn}}

The statement <{spawn}> <{m a}> starts a new thread that executes
method <{m}> with argument <{a}>. <{m}> must be a program counter constant,
and <{a}> is typically a tuple containing
zero or more parameters to be passed to the method.

The default thread-local state of the thread, called <{self}>,
is the empty dictionary by default.  It can be specified by adding
a parameter: <{spawn}> <{m}> $a, e$ specifies that <{e}> should be
the initial value of the thread-local state.

Harmony normally checks that all threads eventually terminate.
If a thread may never terminate, you should spawn it with
\texttt{spawn eternal} <{m a}> to suppress those checks.

@@

\subsection*{\textbf{when}}

@@

Associative operators ($+$, $*$, $|$, $\string&$, $\string^$, <{and}>, \textbf{or})
are interpreted as general $n$-ary operators, and you are allowed to write
$a + b + c$.  However, $a - b - c$ is illegal, as is any combination of operators with an
arity larger than one, such as $a + b < c$.
In such cases you have to add parentheses or brackets to indicate what
the intended evaluation order is, such as $(a + b) < c$.

In almost all expressions, subexpressions are evaluated left to right.
So, <{a[b] + c}>
first evaluates <{a}>, then <{b}> (and then applies <{b}> to <{a}>), and then <{c}>.  The one
exception is the expression <{a if c else b}>, where <{c}> is evaluated
first.  In that expression, only <{a}> or <{b}> is evaluated depending on the value of <{c}>.
In the expression <{a and b and}> $...$, evaluation is left
to right but stops once one of the subexpressions evaluates to <{False}>.
Similarly for \textbf{or}, where evaluation stops once one of the subexpressions
evaluates to <{True}>.
A sequence of comparison operations, such as $a < b < c$, is evaluated left to right
but stops as soon as one of the comparisons fails.

As an aside:
the expression <{a}> \textbf{not in} <{b}> is equivalent to <{not}> (<{a}> \textbf{in} <{b}>).
Harmony generalizes this construct for any pair of a unary (except '$-$') and a binary operator.
In particular, <{a not and b}> is the same as <{not (a and b)}>.  
For those familiar with logic gates, <{not and}> is the equivalent of \texttt{NAND}.
Similarly, <{not =>}> is non-implication.

@@

\begin{itemize}
\item \textbf{choose} \emph{expressions}: Harmony's ability to let the program
choose a value from a set;
\item \emph{thread interleaving}: different threads run pseudo-concurrently
with their instructions interleaved in arbitrary ways;
\item \emph{interrupts}: Harmony programs can set interrupts that can go
off at arbitrary times.
\end{itemize}

@@

Charm has some tricks to significantly reduce the state space to explore.
\begin{itemize}
\item A thread can have local state (program counter, stack,
method variables, and thread-local state variables).
That state is called the \emph{context} of the thread.
The context of a thread cannot be accessed by other
threads, nor by \textbf{invariant} statements.
So, the model checker only interleaves threads at \textbf{Load},
\textbf{Store}, and \textbf{Del} instructions where a thread
interacts with global
variables.
\item Threads are anonymous, and therefore two or more threads can
have the same context.  The state of the model checker therefore maintains
a \emph{bag} (multiset) of contexts rather are than a \emph{set} of contexts.
Thus even if there are hundreds of threads, there may be only tens of
possible context states.
\end{itemize}

@@

\index{logical timestamp}%
Harmony provides support for testing linearizability,
but requires that the programmer
identifies what are known as \emph{linearization points}
in the implementation that indicate exactly \emph{which} sequential
execution the concurrent execution must align with.
\autoref{fig:queuelin} is a copy of \autoref{fig:queueconc} extended with
linearization points.
For each operation (<{get}> and <{put}>), the corresponding
linearization point must occur somewhere between acquiring and releasing
the lock.
%
Each linearization point execution is assigned a logical timestamp.
Logical timestamps are numbered $0, 1, ...$
To do so, we have added a counter (\texttt{time}) to the <{Queue}>.
Method \texttt{\_linpoint} saves the current counter in
\textbf{this}.\texttt{qtime} and increments the counter.
The \textbf{this} dictionary maintains \emph{thread-local state} associated
with the thread (\autoref{ch:harmonymachine})---it contains variables that
can be accessed by any method in the thread.

>>>>

Harmony in fact supports this.
It has an operator <{countLabel l}>,
\index{countLabel operator}%
where <{l}> is the name of the label (in this case, <{cs}>).
The operator returns the number of threads executing at that label.
Method <{countLabel}> only exists for specification purposes---do not
use it in normal code.
If you run the code through Harmony, the assertion should fail because
there is no code yet for safely entering and exiting the critical section.

@@

The two things to note here is the use of the <{choose}>
expression and the <{assert}> statement.  Using <{choose}>,
we are able to express searching through all possible strings of
references without a complicated nested iteration.
Using \texttt{assert}, we are able to express the anomaly we are
looking for.

@@

Peterson's algorithm relies on a \emph{sequential consistent memory model}
and hence the <{sequential}> statement: without it Harmony will
complain about data races.
More precisely, the <{sequential}> statement says that the
program relies on memory load and store instructions operating on
the indicated variables to be performed sequentially, and that this
order should be consistent with the order of operations invoked on each
thread.
The default memory models of C and Java are not sequentially
consistent.
The unfortunately named \textbf{volatile} keyword in Java has
a similar function as Harmony's <{sequential}> keyword.
Like many constructions in Java, its \texttt{volatile} keyword
was borrowed from C and C++.
However, in C and C++, they do \emph{not} provide sequential
consistency, and one cannot implement Peterson's algorithm in
C or C++ directly.

@@

Harmony does not report data races in two cases.
First, using the <{sequential}> statement, you can specify that
concurrent access to the specified variables is intended.
Second, if the accesses are within an atomic statement block, then they
are not considered part of a data race.

@@

\begin{comment}
For the same reason that you should not use atomic
statements in your ordinary synchronization code, you should not use
<{sequential}> statements either.  Neither is supported by regular
programming languages.  They exist in Harmony for the purpose of
defining atomic instructions and for testing.
The statements should only appear in modules that define synchronization
primitives or in test programs.
\end{comment}

@@

This causes the context of the thread to be saved in <{s}> and
the thread to be no longer running.
Another thread can revive the thread using the <{go}>
\index{go statement}%
statement:

@@

Here <{c}> is a context and <{r}> is a Harmony value.
It causes a thread with context <{c}> to be added to the state that has
just executed the <{stop}>
\index{stop expression}%
expression.  The <{stop}> expression returns the value <{r}>.

\autoref{fig:suspension} shows the lock interface using suspension.
It is implemented as follows:
\begin{itemize}
\item A lock maintains both a boolean indicating whether the
lock is currently acquired and a list of contexts of threads that want to
acquire the lock.
\item
<{acquire()}>
\index{acquire}%
acquires the lock if available and suspends the invoking thread if not.
In the latter case, the context of the thread is added to the end of the list of contexts.
Note that <{stop}> is called within an atomic statement block---this is
the only exception to such an atomic statement block running to completion.
While the
thread is running no other threads can run, but when the thread suspends itself
other threads can run.
\item
<{release()}>
\index{release}%
checks to see if any threads are waiting to acquire the lock.
If so, it uses the <{head}> and <{tail}>
methods from the \texttt{list} module (see \autoref{ap:module})
to resume the first thread that got
suspended and to remove its context from the list.
\end{itemize}
Selecting the first thread is a design choice.  Another implementation could
have picked the last one, and yet another implementation could have used
<{choose}> to pick an arbitrary one.  Selecting the first is a common
choice in lock implementations as it prevents starvation.

@@

The reason is that \texttt{queue.get()} changes the state by
acquiring a lock, but the expressions in <{assert}>
statements (or <{invariant}> statements)
are not allowed to change the state.

@@

\autoref{fig:qtest1} shows a program that
checks concurrent <{put}> operations using
<{N}> threads.  Thread \texttt{putter(<{v}>)}
adds <{v}> to the queue.
Thread \texttt{main()} waits until all \texttt{putter}
threads have finished, and then dequeues <{N}>
items, verifying that they form the set
$\{ 0 .. \mathtt{N} - 1 \}$.

@@

While having five different test programs is fine,
it would be nice if we can combine them all in one
test program.
This has to be done with some care, as running the
concurrent test programs simultaneously would
lead to a state space explosion.
Instead, we can leverage the <{choose}>
expression.
\autoref{fig:multitest} shows how one might go
about this.
Harmony will run all the tests, but the
tests do not interfere with one another.

@@

On the left is a Hoare monitor.
When some thread <{p}> is in the bathroom, it can invoke <{signal}>
on some non-empty bedroom associated with a condition that holds.
This select a thread <{q}> in that bedroom.
As a result of invoking <{signal}>, <{p}> goes back into the hall
and <{q}> enters the bathroom.\footnote{Per Brinch Hansen' monitors have
a \textbf{continue} operation instead of <{signal}>.  The difference
is that threads that invoke \textbf{continue} leave the monitor rather
than returning to the hall.}
Importantly, when <{q}> enters the bathroom, the condition that <{q}> was
waiting for is guaranteed to hold.

@@

\autoref{fig:actor} illustrates the actor approach.
There are three client threads that each want to be assigned
a unique identifier from the set $\{ 0, 1, 2 \}$.
Normally one would use a shared 0-initialized counter and a lock.
Each client would acquire the lock, get the value of the counter
and increment it, and release the lock.
Instead, in the actor approach the counter is managed by a
separate server thread.
The server never terminates, so it is spawned with the keyword
<{eternal}> to suppress non-terminating state warnings.
Each client sends a request to the server, consisting in this case
of simply the queue to which the server must send the response.
The server maintains a local, zero-initialized counter variable.
Upon receiving a request, it returns a response with the value of
the counter and increments the counter.  No lock is required.

@@

\autoref{fig:leader} describes the protocol and its test cases in Harmony.
In Harmony, processors can be modeled by threads and there are a variety of ways in
which one can model a network using shared variables.
Here, we model the network as a set of messages.
The \texttt{send} method atomically adds a message to this set.
Messages are tuples with three fields:
(<{dst}>, <{id}>, <{found}>).
<{dst}> is the identifier of the destination processor;
<{id}> is the identifier that is being forwarded; and
<{found}> is a boolean indicating the second phase of the protocol.
The <{receive}>(<{self}>) method looks for all messages
destined for the processor with identifier <{self}>.

@@

Method \texttt{processor}(<{self}>, <{succ}>) is the
code for a processor with identifier <{self}> and successor
<{succ}>.
It starts simply by sending its own identifier to its successor.
The processor then loops until it discovers the identifier of the
leader in the second phase of the protocol.
A processor waits for a message using the Harmony
<{atomically when exists}> statement.
This statement takes the form <{atomically when exists v in s}>: \emph{statement block}
where <{s}> is a set and <{v}> is variable that is bound to an element of <{s}>.
The properties of the statement are as follows:
\begin{itemize}
\item it waits until <{s}> is non-empty;
\item it is executed atomically;
\item <{v}> is selected non-deterministically, like in the <{choose}> operator.
\end{itemize}

@@

\autoref{fig:2pc2} contains the code for transfers and inquiries, as well
as tests.
The <{receive()}> method is used by coordinators in an
<{atomically when exists}>
statement to wait for a response from each bank involved in a transaction.
Argument \texttt{self} is the
identifier of the coordinator and \texttt{sources} is the set of banks.
It returns the empty set if there not yet responses from all banks.
Otherwise it returns a singleton set containing the set of responses,
one for each source.

@@

@@

In Line~22, a processor waits for <{N - F}> messages using
the Harmony <{atomically when exists}> statement.
Since Harmony has to check all possible executions of the protocol, the
<{receive(round, k)}> method returns all \emph{subbags} of messages
for the given round that have size <{k = N - F}>.  The method uses a
dictionary comprehension to filter out all messages for the given <{round}>
and then uses the <{bag.combinations}> method to find all combinations of size <{k}>.
The <{atomically when exists}> statement waits until there is at least one such combination and
then chooses an element, which is bound to the <{quorum}> variable.
The body of the statement is then executed atomically.
This is usually how distributed algorithms are modeled, because they can only interact
through the network.  There is no need to interleave the different processes other
than when messages are delivered.  By executing the body atomically, a lot of
unnecessary interleavings are avoided and this reduces the state space that must
be explored by the model checker significantly.

The body of the <{atomically when exists}> statement contains the core of the algorithm.
Note that <{N - F = 2F + 1}>, so that the number of
messages is guaranteed to be odd.  Also, because there are only 0 and 1 values, there
must exist a majority of zeroes or ones.  Variable <{count[0]}> stores the number
of zeroes and <{count[1]}> stores the number of ones received in the round.
The rules of the algorithm are simple:
\begin{itemize}
\item update <{estimate}> to be the majority value;
\item if the quorum is unanimous, decide the value.
\end{itemize}
After that, proceed with the next round.

@@

A statement can be preceded by the <{atomically}> keyword to
make the statement atomic.
The statement <{atomically: S1; S2; ...}> evaluates statements
<{S1, S2, ...}> atomically.
This means that the statement runs indivisibly---no other thread can
interleave in the atomic statement.  The only exception
to this is if the atomic block executes a <{stop}> expression.
In this case, another thread can run.  When the original thread
is resumed (using a <{go}> statement), it is once again
atomically executing.

@@

The statement <{await b}> is equivalent to <{when b: pass}>.
It is intended to improve readability of your code.

@@

\subsection*{<{for ... in ... [where ...]}>}

@@

\subsection*{<{from ... import}>}

The statement <{from x import a, b, ...}>
imports module <{x}> and makes its constants <{a, b, ...}> also constants
in the current module.  If a module is imported more than once, its code
is only included the first time.  The constants will typically be the
names of methods (program counter constants) within the module.

You can import all constants from a module <{m}>
(including program counter constants) using the statement
<{from m import *}>.
This, however, excludes constants whose names start with the character
\texttt{\_}: those are considered \emph{private} to the module.

\subsection*{<{go}>}

The statement <{go c e}> starts a thread with context <{c}> that
has executed a <{stop}> expression.  The <{stop}> expression
returns value <{e}>.
The same context can be started multiple times, allowing threads to
\emph{fork}.

\subsection*{<{if ... [elif ...]* [else]}>}

Harmony supports <{if}> statements.
In its most basic form, <{if c: S1; S2; ...}> evaluates <{c}>
and executes statements <{S1, S2, ...}> if and only if boolean expression
<{c}> evaluated to true.
Harmony checks that <{c}> is either <{False}> or <{True}>---if neither
is the case, Harmony reports an error.
The statement <{if c: S1, S2, ... else: T1; T2; ...}>
is similar, but executes statements <{T1, T2, ...}>
if and only if <{c}> evaluated to <{False}>.
You can think of <{elif}> $c:$ as shorthand for <{else: if c:}>.

\subsection*{<{import}>}

The statement <{import m1, m2, ...}> imports modules
<{m1, m2, ...}> in that order.
If a module is imported more than once,
its code is only included the first time.
The constants (including method constants) and shared variables declared 
in that module can subsequently be referenced by prepending ``<{m}>.''.
For example, method <{f()}> in imported module <{m}> is invoked
by calling <{m.f()}>.  If you would prefer to invoke it simply as
<{f()}>, then you have to import using the statement
<{from m import f}>.

\subsection*{<{invariant}>}

The statement <{invariant c}> declares that boolean expression <{c}>
is an invariant.  <{c}> is only allowed to read shared variables and
is evaluated atomically after every state change.
If it ever evaluates to <{False}> Harmony reports and error.  Harmony
also reports an error if the expression evaluates to a value other
than <{False}> or <{True}>.

@@

You can introduce new bound variables in a method using the
<{let}> statement.
The statement <{let a = b: S1; S2, ...}> evaluates <{b}>, assigns
the result to read-only variable <{a}>, and evaluates statements
<{S1, S2, ...}>.  <{let}> supports pattern matching, so you can
write <{let x, (y, z) = b: S1; S2, ...}>.  This will only work
if <{b}> is a tuple with two elements, the second of which also being a
tuple with two elements---if not, Harmony will report an error.

<{let}> statements may be nested, such as
<{let a1 = b1 let a2 = b2: S1; S2; ...}>.
Doing so can improve readability by reducing indentation
compared to writing them as separate statements.
Compare the following two examples:

@@

\subsection*{<{pass}>}

The <{pass}> statement does nothing.

\subsection*{<{print}>}

The statement <{print e}> evaluates <{e}>
and adds the result to the print log.  The print log is used to
create an ``external behavior DFA'' for the Harmony program.

\subsection*{<{sequential}>}

In Harmony, shared variable \texttt{Load} and \texttt{Store} operations
are atomic and have \emph{sequential consistency}.
However, Harmony does check for \emph{data races}.  A data race occurs
when two or more threads simultaneously access the same shared variable,
with at least one of the accesses being a \texttt{Store} operation
\emph{outside of an atomic block}.
If so, Harmony will report an error.  This error can be suppressed by
declaring the shared variable as sequential.  In particular, the statement
<{sequential x, y, ...}> specifies that the algorithm assumes
that the given variables have sequential consistency.

@@

\subsection*{<{spawn}>}

The statement <{spawn m a}> starts a new thread that executes
method <{m}> with argument <{a}>. <{m}> must be a program counter constant,
and <{a}> is typically a tuple containing
zero or more parameters to be passed to the method.

The default thread-local state of the thread, called <{self}>,
is the empty dictionary by default.  It can be specified by adding
a parameter: <{spawn m a, e}> specifies that <{e}> should be
the initial value of the thread-local state.

Harmony normally checks that all threads eventually terminate.
If a thread may never terminate, you should spawn it with
<{spawn eternal m a}> to suppress those checks.

@@

\subsection*{<{when}>}

@@

Associative operators ($+$, $*$, $|$, $\string&$, $\string^$, <{and}>, <{or}>)
are interpreted as general $n$-ary operators, and you are allowed to write
$a + b + c$.  However, $a - b - c$ is illegal, as is any combination of operators with an
arity larger than one, such as $a + b < c$.
In such cases you have to add parentheses or brackets to indicate what
the intended evaluation order is, such as $(a + b) < c$.

In almost all expressions, subexpressions are evaluated left to right.
So, <{a[b] + c}>
first evaluates <{a}>, then <{b}> (and then applies <{b}> to <{a}>), and then <{c}>.  The one
exception is the expression <{a if c else b}>, where <{c}> is evaluated
first.  In that expression, only <{a}> or <{b}> is evaluated depending on the value of <{c}>.
In the expression <{a and b and}> $...$, evaluation is left
to right but stops once one of the subexpressions evaluates to <{False}>.
Similarly for <{or}>, where evaluation stops once one of the subexpressions
evaluates to <{True}>.
A sequence of comparison operations, such as $a < b < c$, is evaluated left to right
but stops as soon as one of the comparisons fails.

As an aside:
the expression <{a not in b}> is equivalent to <{not (a in b)}>.
Harmony generalizes this construct for any pair of a unary (except '$-$') and a binary operator.
In particular, <{a not and b}> is the same as <{not (a and b)}>.  
For those familiar with logic gates, <{not and}> is the equivalent of \texttt{NAND}.
Similarly, <{not =>}> is non-implication.

@@

\begin{itemize}
\item <{choose}> \emph{expressions}: Harmony's ability to let the program
choose a value from a set;
\item \emph{thread interleaving}: different threads run pseudo-concurrently
with their instructions interleaved in arbitrary ways;
\item \emph{interrupts}: Harmony programs can set interrupts that can go
off at arbitrary times.
\end{itemize}

@@

Charm has some tricks to significantly reduce the state space to explore.
\begin{itemize}
\item A thread can have local state (program counter, stack,
method variables, and thread-local state variables).
That state is called the \emph{context} of the thread.
The context of a thread cannot be accessed by other
threads, nor by <{invariant}> statements.
So, the model checker only interleaves threads at \textbf{Load},
\textbf{Store}, and \textbf{Del} instructions where a thread
interacts with global
variables.
\item Threads are anonymous, and therefore two or more threads can
have the same context.  The state of the model checker therefore maintains
a \emph{bag} (multiset) of contexts rather are than a \emph{set} of contexts.
Thus even if there are hundreds of threads, there may be only tens of
possible context states.
\end{itemize}

@@

\index{logical timestamp}%
Harmony provides support for testing linearizability,
but requires that the programmer
identifies what are known as \emph{linearization points}
in the implementation that indicate exactly \emph{which} sequential
execution the concurrent execution must align with.
\autoref{fig:queuelin} is a copy of \autoref{fig:queueconc} extended with
linearization points.
For each operation (<{get}> and <{put}>), the corresponding
linearization point must occur somewhere between acquiring and releasing
the lock.
%
Each linearization point execution is assigned a logical timestamp.
Logical timestamps are numbered $0, 1, ...$
To do so, we have added a counter (\texttt{time}) to the <{Queue}>.
Method <{_linpoint}> saves the current counter in
<{this.qtime}> and increments the counter.
The <{this}> dictionary maintains \emph{thread-local state} associated
with the thread (\autoref{ch:harmonymachine})---it contains variables that
can be accessed by any method in the thread.
